---
title: "tidy_data"
author: "Fang Liu"
date: "10/6/2021"
output: html_document
---

Load needed packages and set up
```{r setup}
library(tidyverse)
options(tibble.print_min = 5) #limit the number of lines printed in a tibble 
```

### pivot_longer
```{r load_data}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()

pulse_df
```

Note that the bdi data is spread across four columns, which correspond to 4 observation times. We can solve this problem by using **pivot_longer**. I think this is similar to the transpose function in SAS.
```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_data
```

```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_data
```

Instead of having intermediate calls and data sets. We can try to do everything at once. 
```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  relocate(visit) %>%
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) %>%
  arrange(id, visit)

print(pulse_df, n = 12)
```

In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?

```{r}
litter_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>%
  select(litter_number, gd0_weight, gd18_weight) %>% 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))

litter_data
```


### pivot_wider